Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-16 paper/ms.tex
update ms

\title{Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS}

@@

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is likely induced by dust or gas clumps in orbit at
  the Keplerian corotation radius.  Here, we report new CQVs
  discovered in TESS short-cadence data collected between July~2018
  and Sep~2022.  Our search of \nstarssearched\ K and M dwarfs with
  $T$$<$16 and $d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most
  of these discoveries are new, and they include the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) examples of this class of object currently
  known.  A few objects are outliers: LP 12-502 for instance showed a
  ``dip complex'' with a period and duration that were fixed over
  1,500 cycles, but in detail, this system exhibited between four and
  eight local minima per cycle, and at times also displayed multiple
  periods simultaneously.  We demonstrate that transient corotating
  material is the most viable explanation for this class of object,
  and present evidence supporting the hypothesis that this material is
  sculpted by stellar magnetic fields dominated by high order
  multipoles.  We expect that our sample will facilitate future
  efforts aimed at connecting CQVs to the broader contexts of star,
  disk, and perhaps even exoplanet evolution. 	
\end{abstract}

@@

With the effective temperatures and stellar radii from the SED fit, we
then estimated the stellar masses by interpolating against the PARSEC
isochrones \citep[v1.2S][]{2014MNRAS.444.2525C}.  The need for models
that incorporate some form of correction for young, active M dwarfs is
well-documented
\citep[e.g.][]{2012ApJ...756...47S,2015ApJ...804..146D,2016A&A...593A..99F,2020ApJ...891...29S}.
Plausible explanations for anomalous M dwarf colors and sizes relative
to model predictions include starspot coverage
\citep[e.g.][]{2017ApJ...836..200G}, and potentially incomplete line
lists \citep[e.g.][]{2013A&A...556A..15R}.  In the PARSEC models,
\citet{2014MNRAS.444.2525C} performed an empirical correction to the
temperature--opacity relation drawn from the BT-Settl model
atmospheres, in order to match observed masses and radii of younq
eclipsing binaries.  This is sufficient for our goal of estimating
stellar masses.  Given our observed $\{ \tilde{T}_{\rm eff},
\tilde{M}_\star, \tilde{t} \}$, and approximating their uncertainties
as Gaussian $\sigma_{\tilde{T}_{\rm eff}}$, $\sigma_{\tilde{M}_\star}$
and $\sigma_{\tilde{t}}$, we evaluate a distance $d$ between our
observations and any model PARSEC grid-point $\{ T_{\rm eff}, M_\star,
t \}$ as
\begin{equation}
  d^2 = 
  \left( \frac{\tilde{T}_{\rm eff} - T_{\rm eff}}{\sigma_{\tilde{T}_{\rm eff}}} \right)^2
  +
  \left( \frac{\tilde{M}_{\star} - M_{\star}}{\sigma_{\tilde{M}_{\star}}} \right)^2
  +
  \left( \frac{\tilde{t} - t}{\sigma_{\tilde{t}}} \right)^2,
\end{equation}
in order to assign equal importance to each dimension.  The preferred
model mass is then one that minimizes this distance, and is quoted in
Table~\ref{tab:thetable}.
{\bf lgb todo: real interpolation, rather than nearest neighbor}

@@

		\subfloat{
			\includegraphics[width=0.46\textwidth]{f7d.pdf}
			\includegraphics[width=0.46\textwidth]{f7c.pdf}
		}
	\end{center}
	\vspace{-0.4cm}
	\caption{
		{\bf River plots of the LP 12-502 light curve}, showing (clockwise
		from top-left) Sectors 18-19, 25-26, 53, and 58-59.  A
		two-harmonic sinusoid has been subtracted to highlight the sharp
		dips.  In Sectors 25-26 (cycles 248-315), periods are visible at
		the fundamental period of 18.5611\,hr, as well as at faster and
		slower relative periods (e.g. $\phi$$\approx$0.3 and
		$\phi$$\approx$0.1).  Similar instances of multiple simultaneous
		periods are also visible in Sector 53 and 58-59.  White chunks
		denote missing data.
	}
	\label{fig:lpriver0}
\end{figure*}

@@

\startlongtable
\begin{deluxetable}{rrrrrrlrllrrrrrr}
\tabletypesize{\footnotesize}
%\tabletypesize{\scriptsize}
\tablecaption{Bona fide, candidate, and debunked complex quasiperiodic
  variables from the TESS 2-minute data. {\bf For internal
  review, full versions are available here
  \url{https://www.dropbox.com/scl/fo/twn4s9ckbevf75jqhtoy1/h?rlkey=t5cn8cx2uoc2ptdm9e570kp1b&dl=0}} \label{tab:thetable}}
%\toprule
%TIC & $T$ & $d$ & BP-RP & RUWE & $P$ & assoc & age & teff & rstar & mstar & rcr & qual & Nsectors \\
%\midrule
%\endhead
\startdata
  TIC & $T$ & $d$ & $\bprp$ & RUWE & $P$ & Assoc & Age & $T_{\rm eff}$ & $R_\star$ & $M_\star$ & $R$$_{\rm cr}$ & $P_{\rm sec}$ & Quality & Bin & $N_{\rm sector}$ \\
  -- &   mag & pc &    mag &      -- &        hr &   -- &        Myr &  K                  & $R_\odot$ & $M_\odot$ & $R_\star$         & hr                     & -- & -- & -- \\
\hline
\input{20230613_LGB_RJ_CPV_TABLE_selfnapplied_rounded_data}
\enddata
\tablecomments{This table includes \ngoods\ CQVs (\texttt{Quality}
  flag 1), \nmaybes\ ambiguous CQVs (\texttt{Quality} flag 0), and
  \ndebunked\ impostors (\texttt{Quality} flag -1).  The three-bit
  binarity flag ``Bin'' is for Gaia DR3
  \texttt{radial\_velocity\_error} outliers (bit 1), Gaia DR3
  \texttt{ruwe} outliers (bit 2), and stars with multiple TESS periods
  (bit 3).  The machine-readable version, available online, includes
  additional columns for the Gaia DR2 and DR3 source identifiers, as
  well as the stellar parameter uncertainties.  The age uncertainties
  are typically $\approx \pm 10\%$, but can be asymmetric.  {\bf The
  median temperature, radius, and mass uncertainties are $\pm XXX$\,K,
  $\pm XX$\%, and $\pm YY$\% respectively}.  $N_{\rm sector}$ denotes
  the number of TESS sectors for which {\it any} data are expected to
  be acquired between July~2018 and Oct~2024.  This number is
  generally greater than the number of sectors for which 120-second
  cadence data exist.   Association names and provenance follow
  conventions adopted by \citet{2018ApJ...856...23G}: 
	ABDMG: AB~Doradus moving group \citep{2015MNRAS.454..593B}.
	ARG: Argus \citep{2019ApJ...870...27Z}.
	%APER: $\alpha$~Persei open cluster \citep{2023AJ....166...14B}.
	BPMG: $\beta$~Pic moving group \citep{2015MNRAS.454..593B}.
  CARN: Carina Near moving group \citep{2006ApJ...649L.115Z}.
	COL: Columba \citep{2015MNRAS.454..593B}.
	EPSC: $\epsilon$ Chamaeleontis \citep{2013MNRAS.435.1325M}.
	LCC: Lower Centaurus Crux \citep{2016MNRAS.461..794P}.
	PERI: Pisces-Eridani \citep{2019AJ....158...77C}.
	PLE: Pleiades \citep{2015ApJ...813..108D}.
	TAU: Taurus \citep{1995ApJS..101..117K}.
	THA: Tucana-Horologium assocation \citep{2015MNRAS.454..593B}.
	TWA: TW Hydrae assocation \citep{2015MNRAS.454..593B}.
	UCL: Upper Centaurus Lupus \citep{2016MNRAS.461..794P}.
	USCO: Upper Scorpius \citep{2016MNRAS.461..794P}.
	The ``(?)'' string denotes low-confidence membership.
}
\end{deluxetable}
\clearpage

>>>>

\title{Transient Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS}

@@

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are low-mass
  pre-main-sequence stars with nearly periodic optical modulation.
  The modulation is probably caused by eclipses of dust or gas clumps orbiting at
  the Keplerian corotation radius.  Here, we report new CQVs
  discovered in TESS short-cadence data collected between 2018~July
  and 2022~September.  Our search of \nstarssearched\ K and M dwarfs with
  $T$$<$16 and $d$$<$150\,pc yielded \ngoods\ high-quality CQVs.  Most
  of these discoveries are new, and they include the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) examples of this class of object currently
  known.  A few objects are outliers: LP 12-502 for instance showed a
  ``dip complex'' with a period and duration that were fixed over
  1,500 cycles, but in detail, this system exhibited between four and
  eight local minima per cycle, and at times also displayed multiple
  periods simultaneously.  We demonstrate that transient corotating
  material is the most viable explanation for this class of object,
  and present evidence supporting the hypothesis that this material is
  sculpted by stellar magnetic fields dominated by high order
  multipoles.  We expect that our sample will facilitate future
  efforts aimed at connecting CQVs to the broader contexts of star,
  disk, and perhaps even exoplanet evolution. 	
\end{abstract}

@@

With the effective temperatures and stellar radii from the SED fit, we
then estimated the stellar masses by interpolating against the PARSEC
isochrones \citep[v1.2S][]{2014MNRAS.444.2525C}.  The need for models
that incorporate some form of correction for young, active M dwarfs is
well-documented
\citep[e.g.][]{2012ApJ...756...47S,2015ApJ...804..146D,2016A&A...593A..99F,2020ApJ...891...29S}.
Plausible explanations for anomalous M dwarf colors and sizes relative
to model predictions include starspot coverage
\citep[e.g.][]{2017ApJ...836..200G}, and potentially incomplete line
lists \citep[e.g.][]{2013A&A...556A..15R}.  In the PARSEC models,
\citet{2014MNRAS.444.2525C} performed an empirical correction to the
temperature--opacity relation drawn from the BT-Settl model
atmospheres, in order to match observed masses and radii of younq
eclipsing binaries.  This is sufficient for our goal of estimating
stellar masses.  Given our observed $\{ \tilde{T}_{\rm eff},
\tilde{M}_\star, \tilde{t} \}$, and approximating their uncertainties
as Gaussian $\sigma_{\tilde{T}_{\rm eff}}$, $\sigma_{\tilde{M}_\star}$
and $\sigma_{\tilde{t}}$, we evaluate a distance $d$ between our
observations and any model PARSEC grid-point $\{ T_{\rm eff}, M_\star,
t \}$ as
\begin{equation}
  d^2 = 
  \left( \frac{\tilde{T}_{\rm eff} - T_{\rm eff}}{\sigma_{\tilde{T}_{\rm eff}}} \right)^2
  +
  \left( \frac{\tilde{M}_{\star} - M_{\star}}{\sigma_{\tilde{M}_{\star}}} \right)^2
  +
  \left( \frac{\tilde{t} - t}{\sigma_{\tilde{t}}} \right)^2,
\end{equation}
in order to assign equal importance to each dimension.  The preferred
model mass is then one that minimizes this distance, and is quoted in
Table~\ref{tab:thetable}.  Uncertainties in the masses are propagated
forward from the statistical uncertainties in the radii, temperatures,
and ages.

@@

		\subfloat{
			\includegraphics[width=0.46\textwidth]{f7d.pdf}
			\includegraphics[width=0.46\textwidth]{f7c.pdf}
		}
	\end{center}
	\vspace{-0.4cm}
	\caption{
		{\bf River plots of the LP 12-502 light curve}, showing (clockwise
		from top-left) Sectors 18-19, 25-26, 53, and 58-59.  A
		two-harmonic sinusoid has been subtracted to highlight the sharp
		dips.  In Sectors 25-26 (cycles 248-315), periods are visible at
		the fundamental period of 18.5611\,hr, as well as at faster and
		slower relative periods.  For example, at $\phi$$\approx$0.05 and
		$\phi$$\approx$0.3, the various blue ``dips'' have distinct slopes, rather than being
		exactly vertical.  Multiple simultaneous
		periods are also visible in Sectors 53 and 58-59.  White chunks
		denote missing data.
	}
	\label{fig:lpriver0}
\end{figure*}

@@

\startlongtable
\begin{deluxetable}{rrrrrrlrllrrrrrr}
\tabletypesize{\footnotesize}
%\tabletypesize{\scriptsize}
\tablecaption{Bona fide, candidate, and debunked complex quasiperiodic
  variables from the TESS 2-minute data. {\bf For internal
  review, full versions are available here
  \url{https://www.dropbox.com/scl/fo/twn4s9ckbevf75jqhtoy1/h?rlkey=t5cn8cx2uoc2ptdm9e570kp1b&dl=0}} \label{tab:thetable}}
%\toprule
%TIC & $T$ & $d$ & BP-RP & RUWE & $P$ & assoc & age & teff & rstar & mstar & rcr & qual & Nsectors \\
%\midrule
%\endhead
\startdata
  TIC & $T$ & $d$ & $\bprp$ & RUWE & $P$ & Assoc & Age & $T_{\rm eff}$ & $R_\star$ & $M_\star$ & $R$$_{\rm cr}$ & $P_{\rm sec}$ & Quality & Bin & $N_{\rm sector}$ \\
  -- &   mag & pc &    mag &      -- &        hr &   -- &        Myr &  K                  & $R_\odot$ & $M_\odot$ & $R_\star$         & hr                     & -- & -- & -- \\
\hline
\input{20230613_LGB_RJ_CPV_TABLE_selfnapplied_rounded_data}
\enddata
\tablecomments{This table includes \ngoods\ CQVs (\texttt{Quality}
  flag 1), \nmaybes\ ambiguous CQVs (\texttt{Quality} flag 0), and
  \ndebunked\ impostors (\texttt{Quality} flag -1).  The three-bit
  binarity flag ``Bin'' is for Gaia DR3
  \texttt{radial\_velocity\_error} outliers (bit 1), Gaia DR3
  \texttt{ruwe} outliers (bit 2), and stars with multiple TESS periods
  (bit 3).  The machine-readable version, available online, includes
  additional columns for the Gaia DR2 and DR3 source identifiers, as
  well as the stellar parameter uncertainties.  The age uncertainties
  are typically $\approx \pm 10\%$, but can be asymmetric.  The median
  statistical uncertainties on the temperature, radius, and mass are
  $\pm$50\,K, $\pm$4\% and $\pm$9\% respectively.
  $N_{\rm sector}$ denotes the number of TESS sectors for which {\it
  any} data are expected to be acquired between July~2018 and
  Oct~2024.  This number is generally greater than the number of
  sectors for which 120-second cadence data exist.   Association names
  and provenance follow conventions adopted by
  \citet{2018ApJ...856...23G}: 
	ABDMG: AB~Doradus moving group \citep{2015MNRAS.454..593B}.
	ARG: Argus \citep{2019ApJ...870...27Z}.
	%APER: $\alpha$~Persei open cluster \citep{2023AJ....166...14B}.
	BPMG: $\beta$~Pic moving group \citep{2015MNRAS.454..593B}.
  CARN: Carina Near moving group \citep{2006ApJ...649L.115Z}.
	COL: Columba \citep{2015MNRAS.454..593B}.
	EPSC: $\epsilon$ Chamaeleontis \citep{2013MNRAS.435.1325M}.
	LCC: Lower Centaurus Crux \citep{2016MNRAS.461..794P}.
	PERI: Pisces-Eridani \citep{2019AJ....158...77C}.
	PLE: Pleiades \citep{2015ApJ...813..108D}.
	TAU: Taurus \citep{1995ApJS..101..117K}.
	THA: Tucana-Horologium assocation \citep{2015MNRAS.454..593B}.
	TWA: TW Hydrae assocation \citep{2015MNRAS.454..593B}.
	UCL: Upper Centaurus Lupus \citep{2016MNRAS.461..794P}.
	USCO: Upper Scorpius \citep{2016MNRAS.461..794P}.
	The ``(?)'' string denotes low-confidence membership.
}
\end{deluxetable}
\clearpage
